---
title: "morgancpp: Efficient structure for storing and comparing Morgan fingerprints"
date: "4/27/2020"
output: github_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

## Installation

The package can be installed directly from GitHub:

```{r, echo = TRUE, eval = FALSE}
if(!require(devtools)) install.packages("devtools")
devtools::install_github("labsyspharm/morgancpp")
```

Once installed, the package can be loaded through the standard `library()` interface:

```{r}
library(morgancpp)
```

## Example data

The package works with Morgan fingerprints specified as hexadecimal strings of length 512.
An example set of 100,000 fingerprints is included with the package. It can be loaded as follows:

```{r}
fn <- system.file("examples/example1.txt.gz", package = "morgancpp")
fps <- scan(fn, what = character())
```

```{r, echo = FALSE}
paste0(substr(fps, start = 0, stop = 80), "...")[1:3]
```

## Computing similarity of fingerprints

The tanimoto similarity can be computed directly on hexadecimal strings:

```{r, error = TRUE}
tanimoto(fps[1], fps[1])
tanimoto(fps[1], fps[2])
tanimoto(fps[1], "FFF")
```

However, doing this for a large number of fingerprints will be slow. Instead, we can initialize the c++ data structure to store fingerprints in an efficient manner and query that structure for similarity:

```{r}
m <- MorganFPS$new(fps)
object.size(fps)
m$size()
```

Similarity between fingerprints in the structure can be computed by indexing:

```{r}
m$tanimoto(1, 1)
m$tanimoto(1, 2)
```

The entire similarity profile against all other fingerprints in the collection can be obtained for compounds already in the collection (specified as index `i`) or new external compounds (specified as hexadecimal strings):

```{r}
res1 <- m$tanimoto_all( 1 )       # Compound 1 against all 100,000 fingerprints
```

```{r, echo = FALSE, results = "asis"}
kable(head(res1), label = "res1")
```


```{r}
res2 <- m$tanimoto_ext( fps[1] )  # External compound against all 100,000 fingerprints
```

```{r, echo = FALSE, results = "asis"}
kable(head(res2), label = "res2")
```

## Tresholded searches

Often the similarities between all NxN pairs of fingerprints above a certain
threshold are of interest. The `tanimoto_threshold()` computes these similarities
and outputs them as a data frame.


```{r}
res3 <- m$tanimoto_threshold( 0.9 )
```

```{r, echo = FALSE, results = "asis"}
kable(head(res3), label = "res3")
```

## Saving fingerprints

Fingerprints can be stored on disk in a binary format supporting extremely fast writing
and reading performance using [Zstandard](https://github.com/facebook/zstd)
compression.

```{r, echo = TRUE}
tmp_file <- tempfile()

m$save_file(tmp_file)

m2 <- MorganFPS$new(tmp_file, from_file = TRUE)
m2$size()
```

## Additional documentation

Obtaining additional information about functionality of the package can be done through the standard R interface:

```{r, echo = TRUE, eval = FALSE}
# All available functions
library( help = morgancpp )

# Help for individual functions / data structures
?morgancpp::tanimoto
?morgancpp::MorganFPS

# C++ docstrings providing full signatures of each method
morgancpp::MorganFPS
```

